Temporal ordering and registration of cross-sectional imaging data

### 2014-05-20 image_analysis_paper/pnas_draft2.tex
edits before meeting with Yannis

@@

Here, we show how recently developed dimensionality reduction techniques can successfully address these challenges and reveal developmental trajectories from collections of unregistered and unordered imaging data. 
%
Based on manifold learning algorithms for high-dimensional data (such as images), these techniques have been used for data mining in a wide range of applications, from cryo-EM analysis of individual molecules to face recognition. 
%
Here we adapt these techniques for temporal ordering and registration of imaging data from cross-sectional studies of developmental dynamics, with the main objective of revealing stereotypic developmental trajectories.
%
We demonstrate these techniques using three data sets from a study of {\it Drosophila} embryogenesis, one of the best experimental models for studies of developmental dynamics.
%
The first data set, shown in Figure~\ref{subfig:raw_data1}, is relatively simple and will allow us to both illustrate and validate our methods. 
%
In the second data set, shown in Figure~\ref{subfig:raw_data2}, the dynamics are significantly more complex.
%
Here, we will show that we can recover a trajectory that is qualitatively consistent with current knowledge.
%
The third data set, shown in Figure~\ref{subfig:raw_data3}, consists of data collected from both wild type and mutant embryos.
%
We will demonstrate that these algorithms can not only extract stereotypic dynamics, but also separate data sampled from distinct developmental paths.

@@

@@

\begin{figure}[t]
%\raisebox{8.8cm}{{\figtextfont A}}
%\includegraphics[height=9cm]{raw_data3}
%\begin{minipage}[b]{8.5cm}
\raisebox{6cm}{{\figtextfont B}}
\includegraphics[width=8.4cm]{mut_wt_vdm_embedding}
\raisebox{2cm}{{\figtextfont C}}
\includegraphics[width=8.4cm]{wt_trajectory}
\raisebox{2cm}{{\figtextfont D}}
\includegraphics[width=8.4cm]{mut_trajectory}
%\end{minipage}
\caption{Mutant and wild type {\xfigtextfontit Drosophila}. {\xfigtextfontit (A)} Images of cellularizing {\xfigtextfontit Drosophila} embryos. Each image is of a different embryo in a different rotational orientation. Half of the images are of wild type embryos, and half of the images are of sna null mutant embryos. {\xfigtextfontit(B)} Data from {\xfigtextfontit A}, projected onto the first two vector diffusion maps coordinates. The two distinct trajectories are highlighted in blue and red. {\xfigtextfontit (C)} Average trajectory of the dpERK signal over the blue curve in {\xfigtextfontit B}. {\xfigtextfontit (D)} Average trajectory of the dpERK signal over the red curve in {\xfigtextfontit B}.}
\customlabel{fig:data3}{5}
\customlabel{subfig:raw_data3}{\ref{fig:data3}{\it A}}
\customlabel{subfig:projection_data3}{\ref{fig:data3}{\it B}}
\customlabel{subfig:average_wt_data3}{\ref{fig:data3}{\it C}}
\customlabel{subfig:average_mut_data3}{\ref{fig:data3}{\it D}}
\end{figure}

@@

We have demonstrated how vector diffusion maps, an existing reduction technique, can be used to register and temporally order images from developmental biology studies. 
%
We acknowledge that the tasks of both registration and ordering could have been accomplished using a variety of other techniques.
%
We chose the algorithm outlined here, not because it is specifically tailored to our particular data sets, but rather, because it is sufficiently general and we are confident that it can be applied to a myriad of biological imaging applications.

@@

%\bibliographystyle{pnas}
%\bibliography{background_reading/references,../../references/references}

>>>>

\usepackage{tikz}

@@

Here, we show how recently developed dimensionality reduction techniques can successfully address these challenges and reveal developmental trajectories from collections of unregistered and unordered imaging data. 
%
Image registration has been studied in a variety of contexts \cite{zitova2003image, rowley1998rotation, hajnal2010medical, greenspan1994rotation, zhao2003face}, as have dimensionality reduction techniques \cite{Belkin2003, coifman2005geometric, coifman2006geometric, tenenbaum2000global, roweis2000nonlinear} and their applications \cite{anavy2014blind, trapnell2014dynamics,gupta2008extracting, qiu2011discovering, kafri2013dynamics}.
%
Here, we use a manifold learning algorithm \cite{singer2012vector} for high-dimensional data (such as images) which both factors out symmetries to align images and extracts low-dimensional structure in the imaging data set; these techniques have been used for data mining in a wide range of applications, from cryo-EM analysis of individual molecules to face recognition. 
%
We adapt these techniques for temporal ordering and registration of imaging data from cross-sectional studies of developmental dynamics, with the main objective of revealing stereotypic developmental trajectories.
%
We demonstrate these techniques using three data sets from a study of {\it Drosophila} embryogenesis, one of the best experimental models for studies of developmental dynamics.
%
The first data set, shown in Figure~\ref{subfig:raw_data1}, is relatively simple and will allow us to both illustrate and validate our methods. 
%
In the second data set, shown in Figure~\ref{subfig:raw_data2}, the dynamics are significantly more complex.
%
Here, we will show that we can recover a trajectory that is qualitatively consistent with current knowledge.
%
The third data set consists of data collected from both wild type and mutant embryos.
%
We will demonstrate that these algorithms can not only extract stereotypic dynamics, but also separate data sampled from distinct developmental paths.

@@

We can further use this collection of registered and ordered snapshots to construct a smooth movie of the developmental dynamics. 
%
Using a Gaussian kernel, we calculate the average image for each frame of a movie which (approximately) spans the ordered images. 
%
This movie is included in the {\it SI Appendix}.
%
The dynamics are now apparent.
%
There are artifacts in the movie due to size variation in the samples.
%
Furthermore, the dynamics at the end of the movie are not as smooth, as we have fewer samples in this window. 
%
We would like to note that, although this movie provides a picture of the dynamics, the speed of the movie is not constant with respect to the real time dynamics.

@@

\begin{figure}[t]
\raisebox{5.7cm}{{\figtextfont A}}
\begin{tikzpicture}
     \node at (0,0)
       {\includegraphics[width=8.4cm]{mut_wt_vdm_embedding_background}};
     \node at (0,0)
       {\includegraphics[width=8.4cm]{mut_wt_vdm_embedding}};  
\end{tikzpicture}
\raisebox{0.7cm}{{\figtextfont B}}
\includegraphics[width=8.4cm]{wt_trajectory}
\raisebox{0.7cm}{{\figtextfont C}}
\includegraphics[width=8.4cm]{mut_trajectory}
\caption{Mutant and wild type {\xfigtextfontit Drosophila}.  {\xfigtextfontit(A)} Data collected from wild type and mutant embryos (original data set is shown in {\xfigtextfontit SI Appendix}), projected onto the first two vector diffusion maps coordinates. The two distinct trajectories are highlighted in blue and red. {\xfigtextfontit (B)} dpERK signal at select wild type points in {\xfigtextfontit A} (denoted by bold symbols). {\xfigtextfontit (C)} dpERK signal at select mutant points in {\xfigtextfontit A} (denoted by bold symbols).}
\customlabel{fig:data3}{5}
\customlabel{subfig:projection_data3}{\ref{fig:data3}{\it A}}
\customlabel{subfig:average_wt_data3}{\ref{fig:data3}{\it B}}
\customlabel{subfig:average_mut_data3}{\ref{fig:data3}{\it C}}
\end{figure}

\begin{figure*}[t]
\raisebox{8cm}{{\figtextfont A}}
\includegraphics[height=8.4cm]{wt_trajectory2}
%\raisebox{8cm}{{\figtextfont B}}
\begin{tikzpicture}
     \node at (0,0)
       {\includegraphics[height=8.4cm]{mut_wt_vdm_embedding_background}};
     \node at (0,0)
       {\includegraphics[height=8.4cm]{mut_wt_vdm_embedding}};  
     \node at (-5cm, 3.8cm)
       {{\figtextfont B}};
\end{tikzpicture}
\raisebox{8cm}{{\figtextfont C}}
\includegraphics[height=8.4cm]{mut_trajectory2}
\caption{Mutant and wild type {\xfigtextfontit Drosophila}.  {\xfigtextfontit (A)} dpERK signal at select wild type points in {\xfigtextfontit B} (denoted by bold symbols). {\xfigtextfontit(B)} Data collected from wild type and mutant embryos (original data set is shown in {\xfigtextfontit SI Appendix}), projected onto the first two vector diffusion maps coordinates. The two distinct trajectories are highlighted in blue and red. {\xfigtextfontit (C)} dpERK signal at select mutant points in {\xfigtextfontit B} (denoted by bold symbols).}
\customlabel{fig:data3}{5}
\customlabel{subfig:average_wt_data3}{\ref{fig:data3}{\it A}}
\customlabel{subfig:projection_data3}{\ref{fig:data3}{\it B}}
\customlabel{subfig:average_mut_data3}{\ref{fig:data3}{\it C}}
\end{figure*}

@@

We have demonstrated how vector diffusion maps, an existing reduction technique, can be used to register and temporally order images from developmental biology studies. 
%
We acknowledge that the tasks of both registration and ordering could have been accomplished using a variety of other techniques.
%
We chose the algorithm outlined here, not because it is specifically tailored to our particular data sets, but rather, because it is sufficiently general and we are confident that it can be applied to a myriad of biological imaging applications. 
%
Such algorithms are often tested on databases containing hundreds or thousands of images.
%
Here, we showed that we can successfully recover the underlying dynamics with relatively small data sets.
%
In general, the number of images required to accurately recover the dynamics will be a function of the noise in the data, as well as the dynamic range and complexity.

@@

\bibliographystyle{pnas}
\bibliography{background_reading/references,../../references/references}

### 2014-05-20 image_analysis_paper/si_information.tex
edits before meeting with Yannis

\usepackage{amssymb,amsfonts,amsmath}
\DeclareMathOperator*{\argmin}{arg\,min}

@@

@@

\begin{figure}[h!]
\includegraphics[width=8.4cm, trim=0cm 5.6cm 0cm 0cm, clip]{fig6}
\caption{{\it (A)} Images, stained for membranes (gray) and Dorsal (green) at different levels of development. {\it (B)} Membrane length as a function of time. The points corresponding to the images in {\it A} are indicated with arrows.}
\label{fig:membrane_compare}
\end{figure}

@@

\begin{figure}[h!]
\raisebox{5.2cm}{A}
\includegraphics[width=16.8cm]{data2_PCA_ordered}

@@

\begin{figure}[t]
\includegraphics[width=8.4cm]{fig3}
\caption{{\it (A)} One-dimensional concentration profile on a ring (top), and the corresponding profile on a line (bottom). {\it (B)} Intensity corresponding to the profile in {\it A}. {\it (C)} An ensemble of concentration profiles, each of the form described in {\it A}. Each row in the array corresponds to a single profile. {\it (D)} The profiles in {\it C}, now registered using angular synchronization. {\it (E)} The profiles in {\it D}, now temporally ordered using diffusion maps. {\it (F)} The profiles in {\it C}, registered and temporally ordered in a single step using vector diffusion maps.}
\label{fig:1d_demo}
%\customlabel{fig:1d_demo}{3}
\customlabel{subfig:1d_example}{\ref{fig:1d_demo}{\it A}}
\customlabel{subfig:1d_intensity}{\ref{fig:1d_demo}{\it B}}
\customlabel{subfig:1d_unaligned_unordered}{\ref{fig:1d_demo}{\it C}}
\customlabel{subfig:1d_aligned_unordered}{\ref{fig:1d_demo}{\it D}}
\customlabel{subfig:1d_aligned_ordered}{\ref{fig:1d_demo}{\it E}}
\customlabel{subfig:1d_aligned_ordered_vdm}{\ref{fig:1d_demo}{\it F}}
\end{figure}

@@

\begin{figure*}
\begin{minipage}{0.2\textwidth}
\centering
\includegraphics[width=\textwidth]{sphere_1}\\
\includegraphics[width=\textwidth]{sphere2_1}\\
A
\end{minipage}
%
\begin{minipage}{0.2\textwidth}
\centering
\includegraphics[width=\textwidth]{sphere_2}\\
\includegraphics[width=\textwidth]{sphere2_2}\\
B
\end{minipage}
\begin{minipage}{0.2\textwidth}
\centering
\includegraphics[width=\textwidth]{sphere_3}\\
\includegraphics[width=\textwidth]{sphere2_3}\\
C
\end{minipage}
\begin{minipage}{0.2\textwidth} 
\centering
\includegraphics[width=\textwidth]{sphere_4}\\
\includegraphics[width=\textwidth]{sphere2_4}\\
D
\end{minipage}
\caption{Illustration of how rotations in three dimensions correspond to translations and rotations in two dimensions. {\it (A)} The original image. {\it (B)} Rotation around the x-axis (Euler angle $\alpha$) in three dimensions corresponds to rotation of the image. {\it (C)} Rotation around the y-axis (Euler angle $\beta$) in three dimensions corresponds to horizontal translation. {\it (D)} Rotation around the z-axis (Euler angle $\gamma$) in three dimensions corresponds to vertical translation. The top row shows the three-dimensional spheres, and the bottom row shows the (two-dimensional) surface of the sphere in which we are interested.}
\label{fig:SO3_picture}
\end{figure*}

@@

>>>>

\usepackage{tikz}

\usepackage{float}
\usepackage{amssymb,amsfonts,amsmath}
\DeclareMathOperator*{\argmin}{arg\,min}

@@

\section{Raw data of mutant and wild type embryos}

\begin{figure}[H]
\includegraphics[height=9cm]{raw_data3}
\caption{Images of cellularizing {\it Drosophila} embryos. Each image is of a different embryo in a different rotational orientation. Half of the images are of wild type embryos, and half of the images are of mutant embryos.}
\end{figure}

@@

\begin{figure}[H]
\includegraphics[width=8.4cm, trim=0cm 5.6cm 0cm 0cm, clip]{fig6}
\caption{{\it (A)} Images, stained for membranes (gray) and Dorsal (green) at different levels of development. {\it (B)} Membrane length as a function of time. The points corresponding to the images in {\it A} are indicated with arrows.}
\label{fig:membrane_compare}
\end{figure}

@@

\begin{figure}[H]
\raisebox{5.2cm}{A}
\includegraphics[width=16.8cm]{data2_PCA_ordered}

@@

\begin{figure}[H]
\includegraphics[width=8.4cm]{fig3}
\caption{{\it (A)} One-dimensional concentration profile on a ring (top), and the corresponding profile on a line (bottom). {\it (B)} Intensity corresponding to the profile in {\it A}. {\it (C)} An ensemble of concentration profiles, each of the form described in {\it A}. Each row in the array corresponds to a single profile. {\it (D)} The profiles in {\it C}, now registered using angular synchronization. {\it (E)} The profiles in {\it D}, now temporally ordered using diffusion maps. {\it (F)} The profiles in {\it C}, registered and temporally ordered in a single step using vector diffusion maps.}
\label{fig:1d_demo}
%\customlabel{fig:1d_demo}{3}
\customlabel{subfig:1d_example}{\ref{fig:1d_demo}{\it A}}
\customlabel{subfig:1d_intensity}{\ref{fig:1d_demo}{\it B}}
\customlabel{subfig:1d_unaligned_unordered}{\ref{fig:1d_demo}{\it C}}
\customlabel{subfig:1d_aligned_unordered}{\ref{fig:1d_demo}{\it D}}
\customlabel{subfig:1d_aligned_ordered}{\ref{fig:1d_demo}{\it E}}
\customlabel{subfig:1d_aligned_ordered_vdm}{\ref{fig:1d_demo}{\it F}}
\end{figure}

@@

\begin{figure}[H]
\begin{minipage}{0.2\textwidth}
\centering
\includegraphics[width=\textwidth]{sphere_1}\\
\includegraphics[width=\textwidth]{sphere2_1}\\
A
\end{minipage}
%
\begin{minipage}{0.2\textwidth}
\centering
\includegraphics[width=\textwidth]{sphere_2}\\
\includegraphics[width=\textwidth]{sphere2_2}\\
B
\end{minipage}
\begin{minipage}{0.2\textwidth}
\centering
\includegraphics[width=\textwidth]{sphere_3}\\
\includegraphics[width=\textwidth]{sphere2_3}\\
C
\end{minipage}
\begin{minipage}{0.2\textwidth} 
\centering
\includegraphics[width=\textwidth]{sphere_4}\\
\includegraphics[width=\textwidth]{sphere2_4}\\
D
\end{minipage}
\caption{Illustration of how rotations in three dimensions correspond to translations and rotations in two dimensions. {\it (A)} The original image. {\it (B)} Rotation around the x-axis (Euler angle $\alpha$) in three dimensions corresponds to rotation of the image. {\it (C)} Rotation around the y-axis (Euler angle $\beta$) in three dimensions corresponds to horizontal translation. {\it (D)} Rotation around the z-axis (Euler angle $\gamma$) in three dimensions corresponds to vertical translation. The top row shows the three-dimensional spheres, and the bottom row shows the (two-dimensional) surface of the sphere in which we are interested.}
\label{fig:SO3_picture}
\end{figure}

@@

\section{Eigenvalue Spectra from VDM}

\begin{figure}[H]
\includegraphics[width=5cm]{data1_evals}
\includegraphics[width=5cm]{data2_evals}
\includegraphics[width=5cm]{data3_evals}
\caption{Eigenvalue spectra for three examples.}
\end{figure}

\begin{figure}[H]
\includegraphics[width=5cm]{data1_coord_var}
\includegraphics[width=5cm]{data2_coord_var}
\includegraphics[width=5cm]{data3_coord_var}
\caption{Variance of $\psi_{k,l}$ for three examples. We use the ones that have the largest variance.}
\end{figure}

\begin{figure}[H]
\begin{tikzpicture}
     \node at (0,0)
       {\includegraphics[width=8.4cm]{mut_wt_vdm_embedding_background}};
     \node at (0,0)
       {\includegraphics[width=8.4cm]{mut_wt_vdm_embedding}};  
\end{tikzpicture}
\begin{tikzpicture}
     \node at (0,0)
       {\includegraphics[width=8.4cm]{mut_wt_vdm_embedding2_background}};
     \node at (0,0)
       {\includegraphics[width=8.4cm]{mut_wt_vdm_embedding2}};  
\end{tikzpicture}
\caption{Embedding for third example using two different $\epsilon$ values. The right one (smaller $\epsilon$) has a tighter structure, but the mutants are more removed.}
\end{figure}

\section{FBsPCA}

I computed the FBsPCA basis for the set of images (using only the dpERK signal).
%
I then computed the bispectrum coefficients for each image expanded in this basis, and used these coefficients in a DMAPS calculation.
%
The results are shown below.
%
However, because the bispectrum is not stable to deformations, it does not seem to be able to identify neighboring images.

\begin{figure}[H]
\includegraphics[width=7cm]{FBsPCA_images}
\includegraphics[width=10cm]{FBsPCA_dmaps}
\caption{(Left) Images used for FBsPCA. (Right) First 4 dmaps variables, computed from the bispectrum coefficeints, plotted against the membrane thickness. Note there is no appreciable correlation: each dmaps variable simply selects one image from the data set as ``abnormal''.}
\end{figure}
